---
title: Environmental responses consistently, but not completely, synchronize population dynamics of co-occuring grassland species
layout: 12pt
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{setspace}

## rmarkdown render options
output:
  pdf_document:
    fig_caption: true
    keep_tex: false
fontsize: 12pt
geometry: margin=1in

---
```{r set_up, echo=FALSE, warning=FALSE, message=FALSE}
cbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
               "#CC79A7", "#0072B2", "#D55E00", "#CC79A7")
library(reshape2); library(ggplot2); library(gridExtra)
library(synchrony); library(plyr); library(tidyr); library(xtable)

# This is a new synchrony measure from Gross et al. 2014 (Am. Nat.)
# The function takes a community matrix where columns are species and rows are years (or other time unit)
get_eta <- function(y_matrix){
  n <- ncol(y_matrix)
  tmp_corr <- numeric(n)
  for(i in 1:n){
    y_focal <- y_matrix[,i]
    y_other <- y_matrix[,-i]
    tmp_corr[i] <- cor(y_focal, rowSums(y_other))
  }
  eta_out <- (1/n)*sum(tmp_corr)
  return(eta_out)
}


```


\begin{center}
Andrew T. Tredennick and ... and ...
\end{center}

Introduction
------------
Asynchrony in population dynamics among co-occuring plant species stabilizes ecosystem-level properties like annual biomass production (Hector et al. 2010, Loreau and de Mazancourt 2013, de Mazancourt et al. 2013, Hautier et al. 2014). Species-specific responses to environmental conditions, in particular, can stabilize ecosystem properties even as environmental forcings fluctuate through time. Such asynchrony of species dynamics in response to the environment has the potential to buffer ecosystems subjected to an increasingly variable climate. Thus, the extent to which natural communities exhibit asynchronous dynamics determines, in part, their ability to cope with interannual climate extremes that are projected to become more common.

Theory suggests that community synchrony is driven by a combination of internal and external forces. Internal to the community, demographic stochasticity can induce asynchronous dynamics when species tend to fluctuate independently in a constant environment (Loreau and de Mazancourt 2008). However, species live in non-constant environments that act as strong external drivers of community dynamics. Environmental variability can induce asynchronous dynamics when species' responses to environmental conditions are temporally uncorrelated. When species respond similarly to environmental conditions, community dynamics become more synchronized (Gonzalez and Loreau 2009, Loreau and de Mazancourt 2008).

Given that environmental drivers can either increase or decrease species synchrony, what should be our null expectation of community synchrony? Past studies have assumed the appropriate null is independent species fluctuations (Houlahan PNAS), but this only occurs in the absence of environmental forcing. Species that co-occur but occupy the same trophic level tend to compete for similar abiotic resources, the availability of which varies annually. For example, water is the limiting resource in semi-arid grasslands so annual net primary productivity tends to track annual precipitation (Knapp citation?), regardless of species (CITATION). Taken to the limit, complete similarity among species in terms of resource use and environmentally-mediated growth rates results in communities that exhibit neutral dynamics (Hubbell book). Thus, as suggested by theory (Loreau and de Mazancourt 2008) our neutral expectation is that plant species that co-occur in water-limited grasslands will exhibit perfectly synchronous species dynamics since their growth is strongly driven by precipitation (Knapp again?).

It is impossible to draw inference on the relative contribution of demographic and environmental stochasticity on synchrony based on empirical estimates alone, so our analysis involves two steps.  First, we use exceptional longterm datasets from five North American grasslands (Chu and Adler 2015) to calculate community synchrony. Second, we fit multispecies population models to the data for dominant species to evaluate the relative contribution of species' environmental responses and demographic stochasticity to community synchrony. Sparse data on sub-dominant species constrained model fitting to dominant species that account for ##% - ##% of local plant abundance at their respective sites. Our focus is on temporal trends of per capita growth rates becuase these best reflect the short-term responses of populations to environmental conditions that fluctuate on annual time scales (Loreau and de Mazancourt 2008).

Methods
-------
We built environmentally and demographically stochastic multi-species integral projection models (IPMs) to simulate species' dynamics through time. The IPMs are based on vital rate regressions for survival, growth, and recruitment fit to long-term data (Fig. 1; Chu and Adler 2015). Environmental stochasticity is incorporated by fitting random year effects for the intercept of all regressions and on the plant size effect in the survival and growth regressions. Thus, for a fluctuating environment we can randomly draw year-specific regressions for each time-step of the IPM. Alternatively, to simulate the community in a constant environment we can use the mean regressions for each species. We incorporate demographic stochasticity in the IPM by making survival a binomial process (see Vindenes et al. 2011). We can remove demographic stochaticity by treating survival as a continuous rather than binomial process, as is traditionally done with IPMs (e.g., Rees and Ellner 2014).

To determine the (de)synchronizing effects of demographic stochasticity and species' responses to the environment we simulated communities using the IPM under three scenarios: (1) demographic and environmental stochasticity included, (2) demographic stochasticity removed, and (3) environmental stochasticity removed. We ran simulations for 2,500 time steps and calculated $\phi_r$ using the final 1,000 time steps. For each of our study sites we then calculated $\Delta(\phi)$, the difference between either scenarios 2 or 3 and the unperturbed scenario 1. The sign and magnitude of $\Delta(\phi)$ indicates the (de)synchronizing effect of removing either demographic or environmental stochasticity.


Results
-------
Synchrony of per capita growth rates among species in each community were in the range ## - ## (fill these in after doing all sites) with an average synchrony of ## (Table 1). If our null expectation had been independent fluctuations, then we would conclude that these communities are far from exhibiting asynchronous dynamics that can stabilize annual productivity. However, based on neutral theory we expected species to fluctuate in perfect synchrony. Under that light, our results show that these communities exhibit weak asynchrony since any asynchrony in temporal dynamics drives $\phi_r$ away from zero. 


Given that species are not fluctuating in perfect synchrony, can we conclude that asynchrony in species' responses to the environment is responsible for our observations? To answer this we turn to our simulation results from the IPMs as fit to the observation data. Removing environmental variation consistently increased community asynchrony at each of our study sites (Fig. 2), indicating that these grassland species have similar responses to the environment that synchronize their dynamics. Removing demographic stochasticity, on the other hand, had only modest effects on community synchrony (Fig. 2). Thus, we can conclude, as expected by contemporary theory on species coexistence and community assembly (Loreau and de Mazancourt 2008, HillRisLambers et al. 2013), that these functionally similar perrenial grasses share similar responses to the environment that synchronize their dynamics and, in turn, reduce the temporal stability of biomass production.

Species asynchrony has been heralded as a major stabilizing force in plant communities, one that could buffer the impacts of an increasingly variable climate. However, our results suggest that dominant species in semi-arid grasslands will likely respond to climate fluctuations as a group. This could mean more unstable biomass production in the future if the community does not reorganize. For example, experimental studies show that sub-dominant species can become dominant under novel climate conditions (Hoover et al. 2014, others?). Likewise, if we include a dominant shrub in our analysis of the Idaho dataset it has a stabilizing effect on community biomass because its dynamics are asynchronous to the grasses. Thus, we should anticipate either increasingly unstable biomass production through time as climate extremes become more common, or dramatic community reorganization. In conclusion, our results highlight the importance of not only species diversity, but also functional diversity to ecological stability. 

\newpage

Figures
-------
```{r data_plots, echo=FALSE, fig.height=5, fig.width=8.5, fig.cap="Longterm datasets and the observed per capita growth rates for each year for which contiguous transitions occured.", warning=FALSE, message=FALSE, results='hide', prompt=FALSE}
####
#### 1.1 Kansas ------------------------------------------------------------------
####
# bring in data
site <- "Kansas"
spp_list <- c("BOCU","BOHI","SCSC")
num_spp <- length(spp_list)
ks_data <- data.frame(quad=NA, year=NA, totCover=NA, species=NA)
for(dospp in 1:num_spp){ #loop through species to read in data
  spp_now <- spp_list[dospp]
  quad_file <- paste("../Data/", site,"/",spp_now,"/quadratCover.csv",sep="")
  spp_data <- read.csv(quad_file)
  spp_data$species <- spp_now
  ks_data <- rbind(ks_data, spp_data)
} #end species looping for raw data
ks_data <- ks_data[2:nrow(ks_data),] #remove first NA row

# now exclude some quadrats (from Chengjin's work)
tmp1<-which(ks_data$quad_data=="q25" & (ks_data$year<35 | ks_data$year>62))
tmp2<-which(ks_data$quad_data=="q27")
tmp3<-which(ks_data$quad=="q28")
tmp4<-which(ks_data$quad=="q30")
tmp5<-which(ks_data$quad=="q31" & (ks_data$year<35 | ks_data$year>39))
tmp6<-which(ks_data$quad=="q32" & (ks_data$year<35 | ks_data$year>41))
tmp<-c(tmp1,tmp2,tmp3,tmp4,tmp5,tmp6)
ks_data<-ks_data[-tmp,]

# exclude the records later than 1968, to keep the same random year effect...
ks_data<-subset(ks_data,year<68)

# aggregate the quadrat-level data for average observed cover
# divide totCover by 100 to convert from m2 to percent cover in 1m2 plot
ks_agg <- ddply(ks_data, .(year, species), summarise,
                tot_cover = mean(totCover/100))

# caclulate synchrony and stability
# first stability (mean/sd; Tilman et al. citation) by population and then community
ts_ks <- numeric(num_spp+1)
for(i in 1:num_spp){ #loop through species for population stability
  tmp <- subset(ks_agg, species==spp_list[i])
  ts_ks[i] <- mean(tmp$tot_cover)/sd(tmp$tot_cover)
} #end species looping for stability

# Calculate total cover from the species-level data for each quadrat
ks_sum <- ddply(ks_agg, .(year), summarise,
                all_cover = sum(tot_cover))
ts_ks[num_spp+1] <- mean(ks_sum$all_cover)/sd(ks_sum$all_cover)

# now synchrony of the population fluctuations
ks_mat <- dcast(ks_agg, formula = year~species)
synch_ks <- community.sync(ks_mat[2:4]) #Loreau and de Mazancourt 2008 (Am Nat)
synch_ks_eta <- get_eta(ks_mat[2:4]) #Gross et al. 2014 (Am Nat)

# caclulate observed growth rates
# create lagged data frame to only get observed yearly transitions
lag_df <- ks_mat
lag_df$lagyear <- lag_df$year+1
colnames(lag_df)[2:4] <- paste(colnames(lag_df)[2:4],"_t0", sep="") 
# merge the lag df with observed
merged_df <- merge(ks_mat, lag_df, by.x = "year", by.y="lagyear")
transitions <- nrow(merged_df)
ks_obs_gr <- matrix(nrow=transitions, ncol=num_spp)
for(i in 1:transitions){
  ks_obs_gr[i,] <- as.numeric(log(merged_df[i,2:4]/merged_df[i,6:8]))
}


####
#### 1.2 Idaho ----------------------------------------------------------------------
####
# bring in data
site <- "Idaho"
spp_list <- sort(c("PSSP","HECO","POSE"))
num_spp <- length(spp_list)
id_data <- data.frame(quad=NA, year=NA, totCover=NA, species=NA)
for(dospp in 1:num_spp){ #loop through species to read in data
  spp_now <- spp_list[dospp]
  quad_file <- paste("../Data/", site,"/",spp_now,"/quadratCover.csv",sep="")
  spp_data <- read.csv(quad_file)
  spp_data$species <- spp_now
  id_data <- rbind(id_data, spp_data)
} #end species looping for raw data
id_data <- id_data[2:nrow(id_data),] #remove first NA row

# aggregate the quadrat-level data for average observed cover
# divide totCover by 100 to convert from m2 to percent cover in 1m2 plot
id_agg <- ddply(id_data, .(year, species), summarise,
                tot_cover = mean(totCover/100))

# caclulate synchrony and stability
# first stability (sd/mean) by population and then community
ts_id <- numeric(num_spp+1)
for(i in 1:num_spp){
  tmp <- subset(id_agg, species==spp_list[i])
  ts_id[i] <- mean(tmp$tot_cover)/sd(tmp$tot_cover)
}
id_sum <- ddply(id_agg, .(year), summarise,
                all_cover = sum(tot_cover))
ts_id[num_spp+1] <- mean(id_sum$all_cover)/sd(id_sum$all_cover)


# now synchrony of the population fluctuations
id_mat <- dcast(id_agg, formula = year~species)
synch_id <- community.sync(id_mat[2:4])
synch_id_eta <- get_eta(id_mat[2:4])

# caclulate observed growth rates
# create lagged data frame to only get observed yearly transitions
lag_df <- id_mat
lag_df$lagyear <- lag_df$year+1
colnames(lag_df)[2:4] <- paste(colnames(lag_df)[2:4],"_t0", sep="") 
# merge the lag df with observed (this gives desired 22 year time series)
merged_df <- merge(id_mat, lag_df, by.x = "year", by.y="lagyear")
transitions <- nrow(merged_df)
id_obs_gr <- matrix(nrow=transitions, ncol=num_spp)
for(i in 1:transitions){
  id_obs_gr[i,] <- as.numeric(log(merged_df[i,2:4]/merged_df[i,6:8]))
}


####
#### 2. Some plots -------------------------------
####
#Kansas time series
g_cov_kansas <- ggplot()+
  geom_line(data=ks_agg, aes(x=year, y=tot_cover, color=species))+
  geom_point(data=ks_agg, aes(x=year, y=tot_cover, color=species), size=3)+
  scale_color_manual(values=cbPalette[1:num_spp], name="")+
  xlab("Year (19xx)")+
  ylab("Average Cover (%)")+
  ggtitle("Kansas")+
  guides(color=FALSE)
#Idaho time series
g_cov_idaho <- ggplot()+
  geom_line(data=id_agg, aes(x=year, y=tot_cover, color=species))+
  geom_point(data=id_agg, aes(x=year, y=tot_cover, color=species), size=3)+
  scale_color_manual(values=cbPalette[4:(num_spp+4)], name="")+
  xlab("Year (19xx)")+
  ylab("Average Cover (%)")+
  ggtitle("Idaho")+
  guides(color=FALSE)

colnames(id_obs_gr) <- sort(c("PSSP","HECO","POSE"))
colnames(ks_obs_gr) <- c("BOCU","BOHI","SCSC")
id_obs_growthrate <- as.data.frame(id_obs_gr)
ks_obs_growthrate <- as.data.frame(ks_obs_gr)
id_obs_growthrate$site <- "Idaho"
ks_obs_growthrate$site <- "Kansas"
id_obs_growthrate$year <- c(1:nrow(id_obs_growthrate))
ks_obs_growthrate$year <- c(1:nrow(ks_obs_growthrate))
ks_pgr_frame <- melt(ks_obs_growthrate, id.vars = c("site", "year"))
id_pgr_frame <- melt(id_obs_growthrate, id.vars = c("site", "year"))

#Kansas time series
g_pgr_kansas <- ggplot()+
  geom_line(data=ks_pgr_frame, aes(x=year, y=value, color=variable))+
  geom_point(data=ks_pgr_frame, aes(x=year, y=value, color=variable), size=3)+
  scale_color_manual(values=cbPalette[1:num_spp], name="")+
  xlab("Year")+
  ylab("Per capita growth rate")+
  ggtitle("Kansas")+
  guides(color=FALSE)
#Idaho time series
g_pgr_idaho <- ggplot()+
  geom_line(data=id_pgr_frame, aes(x=year, y=value, color=variable))+
  geom_point(data=id_pgr_frame, aes(x=year, y=value, color=variable), size=3)+
  scale_color_manual(values=cbPalette[4:(num_spp+4)], name="")+
  xlab("Year")+
  ylab("Per capita growth rate")+
  ggtitle("Idaho")+
  guides(color=FALSE)

g_timeseries <- grid.arrange(g_cov_idaho,g_pgr_idaho,g_cov_kansas,g_pgr_kansas, nrow=2, ncol=2)
g_timeseries
```


```{r, echo=FALSE, include=TRUE, fig.height=4, fig.width=5.5, fig.cap="Effect of removing demographic and environmental stochasticity on the synchrony of population dynamics."}
## PLOT FROM 3-25-2015
library(ggplot2)
synch_data <- data.frame(site = rep(c("Idaho", "Kansas"), each=3),
                         simulation = rep(c("Both", "Demo only", "Env only"), times=2),
                         synchrony = c(0.6,0.49,0.58,
                                       0.61,0.51,0.65))
synch_diff <- data.frame(Site = rep(c("Idaho", "Kansas"), each=2),
                         Perturbation = rep(c("Env. Stoch. Removed", "Dem. Stoch. Removed"), times=2),
                         synchrony = c(-0.11, -0.02,
                                       -0.1, 0.04))

ggplot(synch_diff)+
  geom_bar(aes(x=Site, y=synchrony, fill=Perturbation), 
           stat="identity", position='dodge', size=2)+
  geom_hline(aes(yintercept=0))+
  ylab("Perturbed Synchrony - Unperturbed Synchrony")+
  xlab("Site")+
  theme_bw()+
  scale_fill_manual(values=c("grey35", "darkorange"))
#   theme(axis.title.x=element_text(size=14, family="Times New Roman"),
#         axis.title.y=element_text(size=14, family="Times New Roman"),
#         axis.text=element_text(size=12, family="Times New Roman"),
#         legend.text=element_text(size=14, family="Times New Roman"), 
#         legend.title=element_blank())
```


\newpage

Tables
------
```{r synchrony_table, results='asis',message=FALSE, comment=NA, echo=FALSE}
####
#### Get PGR synchrony
####
synch_id_obs <- community.sync(id_obs_gr)
synch_ks_obs <- community.sync(ks_obs_gr)
synch_id_obs_e <- get_eta(id_obs_gr)
synch_ks_obs_e <- get_eta(ks_obs_gr)
synchrony_df <- data.frame(Site = c("Idaho", "Kansas"),
                           Synchrony = c(as.numeric(synch_id_obs[1]),
                                         as.numeric(synch_ks_obs[1])),
                           MeanPairwiseCorrelation = c(as.numeric(synch_id_obs[2]),
                                                         as.numeric(synch_ks_obs[2])))

print.xtable(xtable(synchrony_df, caption = "Community synchrony and mean pairwise correlation between species within a community."),type="latex", comment=FALSE, include.rownames=FALSE, caption.placement="top")
```